Undecidability and the structure of the Turing degrees

### 2018-08-15 paper.tex
further reading, minimal degrees, removed roadmap

\section{Introduction}
Historical overview of the field, motivation for study
\cite{ambos-spies06:_degrees_unsol}
\cite{soare16_turin_comput}
\cite{soare1999history}
\cite{lerman16:_degrees_unsol}
\todo{}
\subsection{Decision Problems}
What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
These questions may seem superfluous, they must be addressed in order to build a
well defined theory of the difficulty of problems. In this paper, we discuss a
specific type of problem, the \emph{decision problem}. Decision problems are
essentially sets of natural numbers, and to solve the problem is to be able to
decide whether any given element is inside the set or not. Because set theory is
commonly used as the foundational system of mathematics, we can convert problems
expressed in terms of more complex mathematical objects into decision problems,
and thus by studying decision problems can build a system that is able to assess
the computability of all mathematical problems. For example, we can express
strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
study problems about languages of such strings with the same tools we use for
decision problems. Going further, we can express the source code of a computer
program as a string over the alphabet $\{0,1\}$, and so can consider questions
about the behavior of computer programs.

@@

The jump operator has many properties that make it convenient for studying the
Turing degrees, which are listed below. Further commentary on the jump operator
and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
\begin{fact}
  The jump operator has many properties that make it convenient for studying the
Turing degrees, some of which are listed below. Further commentary on the jump operator
and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
  \begin{itemize}
  \item $A'$ is computably enumerable in $A$.
  \item $A' \not\leq_T A$.
  \item If $B=_TA$ then $B'=_TA'$.
  \end{itemize}
\end{fact}

@@

\begin{fact}
  \label{fact:not_lat}
  $ \mathcal{D}$ does not form a lattice.
\end{fact}
\begin{rem}
Theorem~\ref{thm:up_sem} and Fact~\ref{fact:not_lat} were first shown in
Kleene's \emph{The Upper Semi-Lattice of Degrees of Recursive
  Unsolvability}~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}. A simpler proof
of Fact~\ref{fact:not_lat} is available in Robert Soare's book \emph{Turing Computability}~\cite[pp.~142--144]{soare16_turin_comput}.
\end{rem}

\begin{thm}
  Existence of minimal degrees
  % Spector 56,Shoenfield 66 (Better)
  \cite{spector56_degrees_recur_unsol}
  \cite{shoenfield66_theor_minim_degrees}
\end{thm}

\begin{thm}
  The c.e. degrees are dense
  \cite{sacks64:_recur_enumer_degrees_dense}
\end{thm}

Homogeneity problems
% Feiner 1970, Shore 1979

Theory of D Undecidable

Theory of D equivalent to to second-order arithmetic \cite{simpson77:_first_order_theor_degrees_recur_unsol}

>>>>

\section{Introduction}
Historical overview of the field, motivation for study
\cite{ambos-spies06:_degrees_unsol}
\cite{soare16_turin_comput}
\cite{soare1999history}
\cite{lerman16:_degrees_unsol}
\todo{}
\subsection{Decision Problems}
What does it mean to \emph{solve a problem}? What is a \emph{problem}, anyways?
These questions may seem superfluous, they must be addressed in order to build a
well defined theory of the difficulty and computability of problems. In this paper, we discuss a
specific type of problem, the \emph{decision problem}. Decision problems are
essentially sets of natural numbers, and to solve the problem is to be able to
decide whether any given element is inside the set or not. Because set theory is
commonly used as the foundational system of mathematics, we can convert problems
expressed in terms of more complex mathematical objects into decision problems,
and thus by studying decision problems can build a system that is able to assess
the computability of all mathematical problems. For example, we can express
strings over the alphabet $\{0,1\}$ as sets of natural numbers, and thus can
study problems about languages of such strings with the same tools we use for
decision problems. Going further, we can express the source code of a computer
program as a string over the alphabet $\{0,1\}$, and so can consider questions
about the behavior of computer programs.

@@

\begin{fact}
  The jump operator has many properties that make it convenient for studying the
Turing degrees, some of which are listed below. Further commentary on the jump operator
and its properties is available in Robert Soare's book \emph{Turing
  Computability}~\cite{soare16_turin_comput}, as well as in Shore and Slaman's
paper \emph{Defining the Turing Jump}~\cite{shore99:_defin_turin_jump}.
  \begin{itemize}
  \item $A'$ is computably enumerable in $A$.
  \item $A' \not\leq_T A$.
  \item If $B=_TA$ then $B'=_TA'$.
  \end{itemize}
\end{fact}

@@

\begin{fact}
  \label{fact:not_lat}
  $ \mathcal{D}$ does not form a lattice.
\end{fact}
\begin{rem}
Theorem~\ref{thm:up_sem} and Fact~\ref{fact:not_lat} were first shown in
Kleene's \emph{The Upper Semi-Lattice of Degrees of Recursive
  Unsolvability}~\cite{kleene54_upper_semi_lattic_degrees_recur_unsol}. A simpler proof
of Fact~\ref{fact:not_lat} is available in Robert Soare's book \emph{Turing Computability}~\cite[pp.~142--144]{soare16_turin_comput}.
\end{rem}
 A final result respecting the structure of $\mathcal{D}$ is the existence of
 \emph{minimal degrees}. A degree $A$ is minimal if $A \neq_T 0$ ($A$ is not
 Turing-decidable), but there does not exist any degree $B\in \mathcal{D}$ such
 that $0 <_T B <_T A$. The existence of such degrees was first shown by Spector
 in 1956~\cite{spector56_degrees_recur_unsol}. An alternative proof is by
 Shoenfield in 1966~\cite{shoenfield66_theor_minim_degrees}.

\section{Further Reading}
The study of the Turing degrees was a very active topic in the late
20\textsuperscript{th} and early 21\textsuperscript{st} centuries, and this
paper has just scratched the surface of the field. \par
Readers particularly interested in the structure of the computably enumerable
degrees might be interested in a result by Sacks in 1964 showing their
density~\cite{sacks64:_recur_enumer_degrees_dense}, while those interested in
\aclp{CFG} or the Turing degrees of problems about them should consult \cite{REEDY197577} and
\cite{ginsburg66:_mathem_theor_contex_languag}
\cite{hopcroft07:_introd_autom_theor_languag_comput}\todo{chapter for cfgs}. For excellent, comprehensive,
and more rigorous textbooks on the study of Turing computability and the Turing
degrees, consider \cite{soare16_turin_comput} \cite{lerman16:_degrees_unsol} or
the classic text \cite{kleene80_introd}.

% \begin{thm}
%   The c.e. degrees are dense
%   \cite{sacks64:_recur_enumer_degrees_dense}
% \end{thm}

% Homogeneity problems
% % Feiner 1970, Shore 1979

% Theory of D Undecidable

% Theory of D equivalent to to second-order arithmetic \cite{simpson77:_first_order_theor_degrees_recur_unsol}
